Designing Types for R, Empirically

### 2020-05-15 paper/main.tex
edits

\newcommand{\NUMFUNCTIONS}{25,000\xspace}  %%% TODO auto-gen
\newcommand{\NUMPACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\PACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\genthat}{genthat\xspace}  %%% TODO auto-gen
\newcommand{\YEARS}{20\xspace} %%TODO fix
\newcommand{\PERCFAILEDASSERTIONS}{2.19\%\xspace}
\newcommand{\PERCFAILEDASSERTIONSWUNDEF}{2.23\%\xspace}
\newcommand{\PERCASSERTIONSUNDEF}{0.04\%\xspace}
\newcommand{\NUMPKGSEVAL}{7485\xspace} % autogenerate
\newcommand{\NUMASSERTIONS}{62,171,573\xspace} %autogenerate
\newcommand{\PROPFUNSFAILEDCHECK}{21.21\%\xspace}
\newcommand{\PROPFUNSFAILEDCHECKNOSTHREE}{15.48\%\xspace} 
\newcommand{\PROPARGSFAILINGASSERTS}{3.53\%\xspace} % autogen....
\newcommand{\PERCSUCCARG}{96.46\%\xspace}
\newcommand{\PERCSUCCFUNNOSTHREE}{84.52\%\xspace}
\newcommand{\PERCSUCCFUNS}{78.79\%\xspace}
\newcommand{\PERCCALLSBADFUNSINTESTS}{5.38\%\xspace}

@@

We implemented \typetracer an automated tool for extracting raw type
signatures from execution traces of R programs. The goal of this tool is to
output a tuple $\langle f, t_1, \dots, t_n, t\rangle$ for each function call
during the execution of a program, where $f$ is an identifier for a
function, $t_i$ are type-level summaries of the arguments and $t$ is a
summary of the return value.

@@

\paragraph{Promises}
The fact that arguments are evaluated lazily, i.e. expression are packing
into promises and evaluated on first access, complicates the information
gathering.  For example, some promises remain un-evaluated, and it would be
erroneous to force them as they may cause side-effects that will affect the
programs execution.  To deal with unevaluated arguments, we make an initial
guess for each argument at fufnction entry.  If the promise is later forced,
we simply update the recorded type for the argument.

@@

\paragraph{Non-local returns}
When a function exits with a longjump there is no return value to speak
of. In order to ensure that call traces are valid when a longjump occurs, we
intercept the unwinding process initiated by the longjump and mimics the
functionality of the functions being exited. When no return value is present
for a call, we record a special \k{jumped} return type.

@@

The output of the \typetracer tool consists of a set of tuples of row types,
each representing a function call in some program's execution. The inference
step consolidate the different tuples corresponding to a particular function
definition across multiple programs and distills them into a single type
signature.  The shape of the infered function signatures is:

@@

\noindent
In other words, we take the union of the types occurring at individual
argument positions rather than an union of function types. Furthermore we
apply some transformation on the types to keep the size of types in check.

@@

\contractr is implemented in 434 R LOC and 2358 C++ LOC. The core
implementation is in C++ to keep the runtime overhead low. It been designed and
tested with GNU R-3.5.0 but it also works with recent versions of R. It has been
hardened with a battery of 400 test cases. We have used it extensively during
the course of this work; firstly, for sanity checking of XXX type signatures
generated for the XXX packages during the development phase, and secondly, for
assessing the quality of these type signatures on XXX packages during the
evaluation phase. We discuss this further in Section~\ref{sec:evaluation}.

@@

For this paper we have selected \CorpusLoadable packages consisting of
\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native
code (C/C++/Fortran). Figure~\ref{fig:corpus} shows all the packages, the
size of the dots reflects the size of the project in terms of both R and
native lines of code\footnote{ Lines of source code reported excludes
  comments and blank lines, counted by \emph{cloc}, \cf
  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
expression code coverage in percents and the y-axis gives the number of
reverse dependencies in log scale. Dotted lines indicate means.  Packages
with over \PackageSizeOutierRnd lines of code are annotated.

>>>>

\newcommand{\NUMFUNCTIONS}{25,000\xspace}  %%% TODO auto-gen
\newcommand{\NUMPACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\PACKAGES}{412\xspace}  %%% TODO auto-gen
\newcommand{\genthat}{genthat\xspace}  %%% TODO auto-gen
\newcommand{\YEARS}{20\xspace} %%TODO fix
\newcommand{\PERCFAILEDASSERTIONS}{2.19\%\xspace}
\newcommand{\PERCFAILEDASSERTIONSWUNDEF}{2.23\%\xspace}
\newcommand{\PERCASSERTIONSUNDEF}{0.04\%\xspace}
\newcommand{\NUMPKGSEVAL}{7485\xspace} % autogenerate
\newcommand{\NUMASSERTIONS}{62,171,573\xspace} %autogenerate
\newcommand{\PROPFUNSFAILEDCHECK}{21.21\%\xspace}
\newcommand{\PROPFUNSFAILEDCHECKNOSTHREE}{15.48\%\xspace} 
\newcommand{\PROPARGSFAILINGASSERTS}{3.53\%\xspace} % autogen....
\newcommand{\PERCSUCCARG}{96.46\%\xspace}
\newcommand{\PERCSUCCFUNNOSTHREE}{84.52\%\xspace}
\newcommand{\PERCSUCCFUNS}{78.79\%\xspace}
\newcommand{\PERCCALLSBADFUNSINTESTS}{5.38\%\xspace}
\newcommand{\TOTNUMSIGS}{21,730\xspace}

@@

We implemented \typetracer, an automated tool for extracting raw type
signatures from execution traces of R programs. The goal of this tool is to
output a tuple $\langle f, t_1, \dots, t_n, t\rangle$ for each function call
during the execution of a program, where $f$ is an identifier for a
function, $t_i$ are type-level summaries of the arguments and $t$ is a
summary of the return value.

@@

\paragraph{Promises}
The fact that arguments are evaluated lazily (in that expression are packed
into promises and evaluated on first access) complicates the information
gathering.  For example, some promises remain un-evaluated, and it would be
erroneous to force them as they may cause side-effects that will affect the
programs execution.  To deal with unevaluated arguments, we make an initial
guess for each argument at function entry.  If the promise is later forced,
we simply update the recorded type for the argument.

@@

\paragraph{Non-local returns}
When a function exits with a longjump there is no return value to speak
of. In order to ensure that call traces are valid when a longjump occurs, we
intercept the unwinding process initiated by the longjump and mimic the
functionality of the functions being exited. When no return value is present
for a call, we record a special \k{jumped} return type.

@@

The output of the \typetracer tool consists of a set of tuples of raw types,
each representing a function call in some program's execution. The inference
step consolidates the different tuples corresponding to a particular function
definition across multiple programs and distills them into a single type
signature.  The shape of the infered function signatures is:

@@

\noindent
In other words, we take the union of the types occurring at individual
argument positions rather than an union of function types. Furthermore, we
apply some transformation on the types to keep the size of types in check.

@@

\contractr is implemented in 434 R LOC and 2358 C++ LOC. The core
implementation is in C++ to keep the runtime overhead low. It been designed and
tested with GNU R-3.5.0 but it also works with recent versions of R. It has been
hardened with a battery of 400 test cases. We have used it extensively during
the course of this work; firstly, for sanity checking of \TOTNUMSIGS type signatures
generated for the XXX packages during the development phase, and secondly, for
assessing the quality of these type signatures on XXX packages during the
evaluation phase. We discuss this further in Section~\ref{sec:evaluation}.

@@

For this paper we have selected \CorpusLoadable packages consisting of
\CorpusRCodeRnd lines of R code and \CorpusNativeCodeRnd lines of native
code (C/C++/Fortran). Figure~\ref{fig:corpus} shows all the packages, the
size of the dots reflects the size of the project in terms of both R and
native lines of code\footnote{ Lines of source code reported excludes
  comments and blank lines, counted by \emph{cloc}, \cf
  \url{https://github.com/AlDanial/cloc}}, the x-axis indicates the
expression code coverage in percent and the y-axis gives the number of
reverse dependencies in log scale. Dotted lines indicate means.  Packages
with over \PackageSizeOutierRnd lines of code are annotated.
